---
title: "Target leakage check: FatFreeW as a predictor of PerBFat1"
format: pdf
---

## Why this file exists

The main analysis (`Multiple linear regression - Body Fat Index.qmd`) retains
`FatFreeW` (fat-free mass) in the final model. `FatFreeW` is not an independent
body measurement. It is derived from the outcome:

$$\text{FatFreeW} = \text{Weight} \times \left(1 - \frac{\text{PerBFat1}}{100}\right)$$

so together with `Weight`, which is also in the dataset, it reconstructs the
outcome almost exactly. Any model containing it reports a fit that cannot be
achieved on a new man with a tape measure, which is the situation the study is
for.

`Density` and `PerBFat2` were already excluded at the start of the main analysis
for exactly this reason. `FatFreeW` belongs in that same exclusion list.

This file quantifies the leak and refits the model without it.

```{r setup, message=FALSE}
d0 <- read.table("../data/fatdat.txt")
names(d0) <- c("ID","PerBFat1","PerBFat2","Density","Age","Weight","Height",
               "AdipInd","FatFreeW","Neck","Chest","Abdomen","Hip","Thigh",
               "Knee","Ankle","Biceps","Forearm","Wrist")
n <- nrow(d0)
```

## 1. The leak, demonstrated arithmetically

```{r}
recon <- 100 * (1 - d0$FatFreeW / d0$Weight)
c(correlation   = cor(recon, d0$PerBFat1),
  mean_abs_diff = mean(abs(recon - d0$PerBFat1)),
  max_abs_diff  = max(abs(recon - d0$PerBFat1)))
```

The correlation is 0.996 and the mean absolute difference is under 0.1
percentage points. The single large discrepancy is a known inconsistent record
in the Penrose data, not a failure of the identity.

```{r}
summary(lm(PerBFat1 ~ FatFreeW + Weight, d0))$r.squared
```

Those two variables alone explain about 96% of the outcome. That number is
arithmetic, not prediction. Its cross-validated error is reported in section 5,
after the helper below is defined.

## 2. Ten-fold cross-validation helper

```{r}
cv_rmse <- function(formula, data, k = 10, seed = 1) {
  set.seed(seed)
  folds <- split(sample(nrow(data)), rep(1:k, length.out = nrow(data)))
  err <- unlist(lapply(folds, function(te) {
    fit  <- lm(formula, data = data[-te, ])
    pred <- predict(fit, newdata = data[te, ])
    pred - data[[all.vars(formula)[1]]][te]
  }))
  sqrt(mean(err^2))
}
```

## 3. The model as originally reported

```{r}
mod_reported <- lm(PerBFat1 ~ poly(FatFreeW, 2) + Height + Chest + Thigh + Wrist, d0)
c(R2 = summary(mod_reported)$r.squared,
  CV_RMSE = cv_rmse(PerBFat1 ~ poly(FatFreeW, 2) + Height + Chest + Thigh + Wrist, d0))
```

## 4. The same selection with FatFreeW excluded from the candidate set

`FatFreeW` is removed before stepwise selection runs, not dropped afterwards.
Dropping it at the end would already have let it shape the selection path.

```{r}
d_clean <- subset(d0, select = -c(ID, PerBFat2, Density, FatFreeW))
mod_full  <- lm(PerBFat1 ~ ., d_clean)
mod_clean <- step(mod_full, k = log(n), trace = 0)   # BIC, explanatory goal
formula(mod_clean)
c(R2 = summary(mod_clean)$r.squared,
  CV_RMSE = cv_rmse(formula(mod_clean), d_clean))
```

## 5. Reference points

```{r}
c(abdomen_only_R2 = summary(lm(PerBFat1 ~ Abdomen, d0))$r.squared,
  abdomen_only_CV = cv_rmse(PerBFat1 ~ Abdomen, d0),
  leak_only_CV    = cv_rmse(PerBFat1 ~ FatFreeW + Weight, d0),
  sd_outcome      = sd(d0$PerBFat1))
```

## Conclusion

Excluding `FatFreeW` costs roughly 0.13 of $R^2$ and raises cross-validated
error modestly. The remaining model uses only measurements that can be taken
with a tape measure and a scale, which is what the study set out to do. The
corrected figures are the ones reported in the README.
